UDC 911.3

### 2024-03-05 legal-companies-concentration-and-availability/Paper.Rmd
Proofreading, fitting into 5 pages

---
title: "UDC 911.3"
subtitle: "Does Concentration of Legal Companies Increases Accessibility of Legal Services: Empirical Evidence from Russian Cities"
output:
  word_document:
    reference: reference.docx
bibliography: biblio.bib
csl: biblio-style.csl
---

@@

The paper is published for the first time.

@@

**Abstract:** a link between concentration of legal companies and availability of legal services is examined. For the analysis, a dataset on the location of small and medium-sized legal companies and their revenue, expenditure and number of employees was made based on Federal Tax Service open data. The concentration is calculated as a ratio of the number of employees of legal companies in a city and a total number of employees of all legal companies in Russia, and the number of employees of legal companies per 100k population was used to measure the availability. Data visualization and linear regression were used to estimate the link. Positive linear relationship between the concentration and availability is found. However, there are various ways to reach good availability of legal services.

@@

For a long time cities have been thought as centers of regions, providing other settlements with goods and services [@Christaller1966]. It is known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019]. Concentration provides economic benefits for the companies and other becomes self-sustaining. However, the increase in the number of firms do not necessarily leads to the corresponding growth in the accessibility of services they provide. An issue of access to various services also has been of great importance for decades [@Blacksell1990; @Keleher1996], because it is directly connected to the quality of life and human rights. There is also a research on the links between market properties and services availability [@Lu2020]. Typically, the literature focuses on few kinds of services, e. g. medical [@Joseph1982], but other professional services also worth attention, e. g. legal services [@Blacksell1988; @Patel2008].

In this paper, the link between the concentration of legal companies in Russian cities and availability of these services is empirically examined. The analysis is based on information about the location of small and medium-sized legal businesses in Russia obtained from open data published by Federal Tax Service (FTS). A main research question is how the availability of legal professionals depends on the concentration of legal companies. This relationship is tested with respect to city population and relative rank in the region (constituent entity) of Russia. An auxiliary problem is to find (if any) geographical patterns in the spatial distribution of cities by legal services availability.

**Data.** Company-wise disaggregated geo-referenced dataset on small and medium-sized legal companies in Russia was prepared and used in this research. The dataset was created with Python CLI tool *ru-smb-companies* developed by the author of this paper. This tool downloads, extracts, aggregates and geocodes large archived XML dumps from the state registry of small and medium-sized businesses (SMBs) distributed by FTS, producing a tiny CSV dataset. To select legal companies, a filtering by main activity code equal to 69.10 was applied. In addition, the dataset was filtered by the following conditions: year is 2021, company is an organization rather than a sole entrepreneur, company has declared non-zero revenue for the target year, company's address of registration is in a city rather than in a rural area. Also, during the exploratory data analysis, four cities (namely, Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were marked as outliers by the count of companies or employees and excluded from the data.

**Methods.** *Concentration of legal companies* in cities was estimated as a number of employees of legal companies in a given city divided by the total number of employees in legal companies in all the cities. The formula was derived and adapted from the Atlas of Economic Specialization of Russian Regions [@Atlas2021]. *Availability of legal services* was measured as the number of employees of legal companies in a city divided by the city's population. This formula is an adapted version of the legal market saturation metric used in the analysis of Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016]. To categorize the cities by population, the classification of Lappo [@Lappo1997] was used. Data visualization, including mapping, was used as the main method, and linear regression as a helper method. Technically, the analysis was performed with R programming language and additional packages. The source code for this paper is available at the public repository: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability

**Results and discussion.** The relationship between the concentration of legal companies and availability of legal services (as measured with the metrics proposed in Methods section) is shown in Figure 1. Apart from the general trend, the groups by city population and city rank within the region are also displayed. The overall connection between the concentration and availability is linear. The results of linear modelling proves it: the link between the two variables can be described with the formula *`r ca_model_formula`*, where both the intercept and slope are highly significant *(p < 0.001)*, and the explained variance is relatively high *(R^2^ = `r round(ca_model_summary$adj.r.squared, 2)`)*.

Looking at the regression line common to all the data points, it is easy to notice that despite the general linearity and good linear modelling results, the total distribution is also fan-shaped, that is, the variance of error of the linear model declines as the both dependent and independent variables grow. However, this is mostly the result of several distinct population size groups of cities present in the data. Within each of these groups, the assumptions of linear models are far better satisfied. The link between concentration and availability is similar in all city population groups, and the regression lines are nearly parallel, meaning that the slope coefficient is equal, and only the intercept differs. Also, it can be concluded that in general, the larger is the city, the bigger is the concentration and the bigger is the availability, even though the relationship is relatively weak.

The first (top rank) cities in the regions usually have better availability. This observation is partially explained by the considerably big size of such cities, despite several exceptions. Also, the top rank of the city within the region does not always lead to the highest possible availability of legal services.

Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration and one of the highest levels of availability of legal services. Moscow accounts for almost 1/3, and Saint Petersburg for about 10% of all employees of legal companies. The relative position of the corresponding data points (compared to the main regression lines) implies that there is some sort of “availability saturation”, that is, the level of concentration where the marginal growth of concentration has little effect on availability.

So, high concentration of legal companies, big population, and the status of first city within the region are positively associated with the availability of legal services. On the contrary, the larger is the city, the more legal companies it must have to satisfy the people's needs. As a result, there are different ways to reach high availability of legal services that range from being the business and service center with high concentration that leads to good availability to being a small yet well-served city where a small flock of legal firms satisfy the needs of the local community.

```{r fig-1-concentration-and-accessibility, fig.cap="Figure 1. The link between the concentration of legal companies and the availability of legal services. In addition, city population group and relative city rank within the region are marked. The formulas for the metrics are described in the Methods section. Both X and Y axes are log10-scaled. Dotted regression lines represent the linear fits within city population groups, and the solid gray line represend the overall linear fit", fig.height=5}
ca_plot
```

The spatial distribution of cities by legal services availability is shown on the map below (Figure 2). Regional centers have high relative level of availability. Moscow and its satellite cities form the most prominant cluster of good access to legal services. Other notable centers with high availability are Krasnodar and Stavropol at the southern part of Russia's East, Yekaterinburg in the Urals, Tuymen in Western Siberia. There are regions with low accessibility of legal services, for example, bordering regions in the South of Siberia, Komi and Kalmyk republics.

```{r fig-2-map, fig.cap="Figure 2. Legal service availability in Russian cities on the country's map", fig.height=5}
spatial_plot
```

**Conclusion.** The concentration of legal companies in cities has a positive linear relationship with the legal services availability. The similar linear trend is observed in all the city population groups. The largest cities of regions usually have relatively high availability of legal services, and the same is true for the cities with population above 1 million. Moscow and Saint Petersburg have extremely high concentration and high availability. Probably there is a saturation limit after which the increase of concentration do not have positive impact on availability. Empirical data suggests several paths for reaching high levels of legal services availability, where increasing concentration of legal companies in a city is just one of the options, and cities with relatively small concentration can achieve the level of accessibility similar to the largest ones. Moscow and its surrounding area forms a remarkable cluster of high legal services availability. As for other places, good levels of access to legal services are noticed in Saint Petersburg, Yekaterinburg, Krasnodar, Tyumen, Stavropol.

>>>>

---
title: "UDC 911.3"
subtitle: "Concentration of Legal Companies and Accessibility of Legal Services: Empirical Evidence from Russian Cities"
output:
  word_document:
    reference: reference.docx
bibliography: biblio.bib
csl: biblio-style.csl
---

@@

The paper is being published for the first time.

@@

**Abstract:** the article examines the relationship between the concentration of legal companies and the availability of legal services. To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia. The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies. The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia. Availability is measured as the number of employees of legal companies per 100,000 population. Data visualization and linear regression were utilized to estimate the hypothesized relationship. The study showed a positive linear relationship between concentration and availability. However, there are multiple approaches to achieving good availability of legal services.

@@

Cities have traditionally been viewed as regional centers that provide goods and services to smaller surrounding settlements [@Christaller1966]. It is well-known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019]. Concentration of businesses provides economic benefits for the companies and thus is likely to become self-sustaining. However, an increase in the number of firms does not necessarily lead to a corresponding growth in the availability of services they provide. The issue of access to various services has been of great importance for decades [@Blacksell1990; @Keleher1996], because it is directly connected to the quality of life and human rights. Research has also been conducted on the links between market properties and service availability [@Lu2020]. Typically, the literature focuses on medical services [@Joseph1982], but other professional services are also worth attention, such as legal services [@Blacksell1988; @Patel2008].

This paper empirically examines link between the concentration of legal companies in Russian cities and the availability of legal services. The analysis is based on information about the location of small and medium-sized legal businesses in Russia obtained from open data published by the Federal Tax Service (FTS). The main research question is how the availability of legal professionals is affected by the concentration of legal companies. Additionally, the study aims to identify any geographical patterns in the spatial distribution of cities based on legal services availability.

**Data.** In this research, a company-wise geo-referenced dataset on small and medium-sized legal companies in Russia was used. The dataset was created using the Python CLI tool *ru-smb-companies* developed by the author of this paper[^1]. The tool downloads, extracts, aggregates and geocodes large archived XML dumps from the state registry of small and medium-sized businesses (SMBs) distributed by FTS, producing a compact CSV dataset. To select legal companies, a filtering by main activity code equal to 69.10 was applied. Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue for the target year, and the company's address of registration is in a city rather than in a rural area. During the exploratory data analysis, four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were identified as outliers based on the number of companies or employees per 100,000 population and were subsequently excluded from the data.

[^1]: https://github.com/PavelSyomin/russian-smb-companies

**Methods.** The *concentration of legal companies* in cities was calculated by dividing the number of employees of legal companies in a given city by the total number of employees of legal companies in all cities. This formula was derived and adapted from the Atlas of Economic Specialization of Russian Regions [@Atlas2021]. The *availability of legal services* was measured by dividing the number of employees of legal companies in a city divided by the city's population. This formula is a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016]. To categorize the cities by population, the classification system of Lappo [@Lappo1997] was used. The main method employed was data visualization, including mapping, with linear regression used as a supplementary method. The analysis was performed using the R programming language and additional packages. The source code for this paper can be found at the public repository[^2].

[^2]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability

**Results and discussion.** Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in Methods section. The plot displays the groups by city population and city rank within the region, in addition to the general trend. The overall connection between the concentration and availability appears to be linear. The linear model results demonstrate that the relationship between the two variables can be described by the formula *`r ca_model_formula`*, where both the intercept and slope are highly significant *(p < 0.001)*, and the explained variance is relatively high *(R^2^ = `r round(ca_model_summary$adj.r.squared, 2)`)*.

Examining the regression line common to all data points, we see that while the linearity is generally good and the linear modeling results are satisfactory, the total distribution is fan-shaped. This means that the variance of error of the linear model decreases as both the dependent and independent variables increase. However, this is mainly due to the presence of several distinct population size groups of cities in the data. Within each of these groups, linear models satisfy the assumptions better. The link between concentration and availability is similar across all city population groups, and the regression lines are nearly parallel, indicating equal slope coefficients and differing only in intercepts. Additionally, it can be concluded that, in general, larger cities have higher concentrations and availability, although the relationship is relatively weak.

```{r fig-1-concentration-and-accessibility, fig.cap="Figure 1. The relationship between the concentration of legal companies and the availability of legal services. City population group and relative city rank within the region are marked. The formulas used to calculate the metrics are described in the Methods section. Both the X and Y axes are log10-scaled. Dotted regression lines represent the linear fits within city population groups, while the solid gray line represent the overall linear fit", fig.height=4}
ca_plot
```

In general, cities with the top rank by the population in their respective regions tend to have better availability of legal services. One possible reason for this is a considerably big size of such cities.

Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability. Moscow alone accounts for almost one-third of all legal companies employees, and Saint Petersburg accounts for about 10%. The position of the corresponding data points relative to the main regression lines suggests that there is an “availability saturation” level, where further concentration has little effect on availability.

A high concentration of legal companies, a large population, and the city's top rank by population within a region are positively associated with the availability of legal services. Conversely, larger cities require more legal companies to meet the needs of their residents. As a result, there are various ways to achieve high availability of legal services, ranging from being a business and service center with a high concentration of legal firms to being a small yet well-served city where a small number of legal firms meet the needs of the local community.

Figure 2 displays the spatial distribution of cities based on the availability of legal services. Regional centers exhibit a high relative level of availability. The most prominent cluster of good access to legal services is formed by Moscow and its satellite cities. Other notable centers with high availability include Krasnodar and Stavropol in the southern part of Russia's West, Yekaterinburg in the Urals, and Tuymen in Western Siberia. There are regions with limited access to legal services, such as the bordering regions in the south of Siberia, as well as the Komi and Kalmyk Republics.

```{r fig-2-map, fig.cap="Figure 2. Legal services availability in Russian cities on the country's map", fig.height=4}
spatial_plot
```

**Conclusion.** The concentration of legal companies in cities has a positive linear relationship with the availability of legal services. This trend is observed in all the city population groups. The largest cities in regions typically have relatively high availability of legal services, as do cities with population above 1 million. Moscow and Saint Petersburg are characterized by the highest concentration and high availability, with Moscow alone accounting for almost one-third of all legal companies employees in the country. Empirical data suggests that there may be a saturation limit beyond which the increase of concentration do not have a substantial impact on availability. There are multiple paths to achieving high levels of legal services availability, with increasing concentration of legal companies in a city being just one option. Even cities with relatively low concentration can achieve similar levels of availability as larger ones. Moscow and its region are a cluster of areas with high legal services availability. Similarly, Saint Petersburg, Yekaterinburg, Krasnodar, Tyumen, and Stavropol also have good access to legal services.
